---
title: "K-Means Clustering"
subtitle: "MUSA 500, Homework #5"
author: "Minwook Kang, Nissim Lebovits, and Ann Zhang"
date: today
format: 
  html:
    toc: true
    theme: lux
    code-fold: true
    code-summary: "Show the code"
editor: visual
execute:
  cache: true
  warning: false
  error: false
  messages: false
---

## Introduction

**1. Describe the data set and indicate what the purpose of this assignment is. That is, how can k-means clustering help you look at the data, and what kinds of questions can you answer?**

```{r data and package import}

#install.packages(c("NbClust","flexclust"))

library(NbClust)
library(flexclust)
library(dplyr)
library(kableExtra)
library(sf)

```

## Methods

**1. How does the K-means algorithm work? Describe the steps in your own words.**

**2. What are some of the limitations of the algorithm?**

**3. What are some other clustering algorithms, and might they be more appropriate here?**

## Results

**1. Present and describe the results from the NbClust command and the scree plot.** **a. What's the optimal number of clusters that you should choose based on the output?**

The flee shows SSE(Sum of sqaure errors) of each number of clusters from 1 to 20. In the graph below, there is a distinct drop of SSE, when cluster moves from 1 to 3. After four clusters, rate of decreasing drops off, suggesting that 3-cluster solution is likely to be a good option to our data.

```{r scree plot and criteria}

regression <- read.csv("https://github.com/nlebovits/musa-500-hmwk-5/raw/main/data/RegressionData.csv") 
data <- read.csv("https://github.com/nlebovits/musa-500-hmwk-5/raw/main/data/RegressionData.csv") 
data <- data %>% select(-POLY_ID, -AREAKEY) 
data <- scale(data)

data_see <- (nrow(data)-1)*sum(apply(data,2,var))

#nrow(data)-1
#apply(data,2,var)

kmeans(data, centers=2)
for (i in 2:20) data_see[i] <- sum(kmeans(data, centers=i)$withinss)

#data_see

plot(1:20, data_see, type="b", xlab="Number of Clusters",
     ylab="Within groups sum of squares")


set.seed(100)
nc <- NbClust(data, min.nc=2, max.nc=15, method="kmeans", index="all")

table(nc$Best.n[1,])

par(mfrow=c(1,1)) 
barplot(table(nc$Best.n[1,]),
        xlab="Numer of Clusters", ylab="Number of Criteria",
        main="Number of Clusters Chosen by 26 Criteria")

```

**2. Present and describe the table produced by the aggregate command showing the mean values of the MEDHVAL, MEDHHINC, PCTBACHMOR, PCTSINGLES, and PCTVACANT in each cluster.** **a. Here, state whether the cluster solution makes sense and if so, come up with descriptive names for each of the resulting clusters.**

```{r aggregated value}

set.seed(1234)
fit.km <- kmeans(data, 3, nstart=1000)

fit.km$size
round(fit.km$centers, 2)
fit.km$cluster

cluster_info = as.data.frame(fit.km$cluster)

table <- cbind(round(aggregate(data, by=list(cluster=fit.km$cluster), mean),1),fit.km$size)
 
colnames(table)[1] = c("CLUSTER")
colnames(table)[8] = c("SIZE")

table <- table[, c(1, 8, 2, 3, 4, 5, 6, 7)]
  
table %>%
  kbl() %>%
  kable_paper("hover", full_width = F)

```

**3. Look up the syntax for the write.csv command and export the table containing the cluster ID of each observation into a .csv file (i.e., the .csv file should contain the variable that indicates which K-means cluster each observation falls into).**

**4.	In ArcGIS, join the .csv file that you exported from R to the RegressionData.shp by the field POLY_ID, and create a map showing the spatial distribution of the clusters.**
**a.	State whether observations falling into the same K-means clusters also tend to cluster in space. That is, does the K-means cluster membership variable seem to be spatially autocorrelated?**
**i.	You should not use Moran’s I here, because the cluster membership variable is categorical.**
**b.	Does looking at the map yield any additional insight into the patterns you observe with the K-means analysis? Does it have any impact on how you might name the clusters?**

```{r scree plot}

Regression_cluster <- cbind(regression, cluster_info)

regression_shp <- st_read("C:/Users/vestalk/Desktop/00_Upenn/20.Fall/03.MUSA 5000 Spatial Statistics and Data Analysis/Assignment/HW 5/RegressionData.shp") 

regression_shp <- left_join(regression_shp, Regression_cluster, by = "POLY_ID") %>% st_as_sf()

regression_cluster <- regression_shp %>% mutate(cluster = fit.km$cluster) %>% dplyr::select(cluster)

plot(regression_cluster)

```

## Discussion

**1.	Briefly describe any patterns that you observe. Any surprising findings?**
